---
title: "Google Analytics Capstone: Cyclistic Bikes R Documentation"
author: "Hannah J. Kissinger"
date: "2023-02-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
This article contains the R Script for the Google Analytics Capstone: Cyclistic Bikeshare. If you are interested in learning more about the derived insights click [here](link here)!


Make sure to set appropriate working directory. Then load the required packages and data from .csv files. 
```{r Load data, echo=TRUE, message=FALSE, warning=FALSE}

#Load packages
library(tidyverse)
library(plyr)

setwd("path/goes/here")

#Load data
Divvy_Trips_2019_Q2 <- read_csv("Divvy_Trips_2019_Q2.csv")
Divvy_Trips_2019_Q3 <- read_csv("Divvy_Trips_2019_Q3.csv")
Divvy_Trips_2019_Q4 <- read_csv("Divvy_Trips_2019_Q4.csv")
Divvy_Trips_2020_Q1 <- read_csv("Divvy_Trips_2020_Q1.csv")

```

Compare the tables' column names.
```{r Column names}
colnames(Divvy_Trips_2019_Q2)
colnames(Divvy_Trips_2019_Q3)
colnames(Divvy_Trips_2019_Q4)
colnames(Divvy_Trips_2020_Q1)
```

Identify the most common column names or use stakeholder's definitions. Change column names to match the chosen identifier. 

Decision: Changed column names from 2019 Q2 and 2020 Q1 to match 2019 Q3 and 2019 Q4.
```{r Rename columns, echo=TRUE, message=FALSE, warning=FALSE}
#Standardize column names in 2019 Q2 and 2020 Q1 dataframes
DT_2019_Q2 <- plyr::rename(Divvy_Trips_2019_Q2, c('01 - Rental Details Rental ID' = 'trip_id',
         '01 - Rental Details Duration In Seconds Uncapped' = 'tripduration',
         '01 - Rental Details Local Start Time' = 'start_time',
         '01 - Rental Details Local End Time' = 'end_time',
         '03 - Rental Start Station ID' = 'from_station_id',
         '03 - Rental Start Station Name' = 'from_station_name',
         '02 - Rental End Station ID' = 'to_station_id',
         '02 - Rental End Station Name' = 'to_station_name',
         'User Type' = 'usertype'))

DT_2020_Q1 <- plyr::rename(Divvy_Trips_2020_Q1, c('ride_id' = 'trip_id',
         'start_station_id' = 'from_station_id',
         'started_at' = 'start_time',
         'ended_at' = 'end_time',
         'start_station_name' = 'from_station_name',
         'end_station_id' = 'to_station_id',
         'end_station_name' = 'to_station_name',
         'member_casual' = 'usertype'))

#Create dataframes for 2019 Q3 and 2019 Q4
DT_2019_Q3 <- Divvy_Trips_2019_Q3
DT_2019_Q4 <- Divvy_Trips_2019_Q4
```

Add the missing trip duration calculated column to the 2020 Q1 dataframe.
```{r Add calculated trip duration, echo=TRUE, message=FALSE, warning=FALSE}
#Load package
library(lubridate)

#Add trip duration column
DT_2020_Q1 <- DT_2020_Q1 %>% 
  mutate(tripduration = as.numeric(difftime(end_time, start_time, units = 'secs')))
```

Select relevant columns for analysis. 
```{r Select columns, echo=TRUE, message=FALSE, warning=FALSE}
#Check column names
colnames(DT_2019_Q3)

#Select columns
DT_2019_Q2_df <- DT_2019_Q2 %>% 
  select('trip_id', 'start_time', 'end_time', 'tripduration', 'from_station_id', 'from_station_name', 'to_station_id', 'to_station_name', 'usertype')

DT_2019_Q3_df <- DT_2019_Q3 %>% 
  select('trip_id', 'start_time', 'end_time', 'tripduration', 'from_station_id', 'from_station_name', 'to_station_id', 'to_station_name', 'usertype')

DT_2019_Q4_df <- DT_2019_Q4 %>% 
  select('trip_id', 'start_time', 'end_time', 'tripduration', 'from_station_id', 'from_station_name', 'to_station_id', 'to_station_name', 'usertype')

DT_2020_Q1_df <- DT_2020_Q1 %>% 
  select('trip_id', 'start_time', 'end_time', 'tripduration', 'from_station_id', 'from_station_name', 'to_station_id', 'to_station_name', 'usertype')

```

Add column for quarter-year to each dataframe.
```{r Add quarter-year}
DT_2019_Q2_df <- DT_2019_Q2_df %>% 
  mutate(quarter_year = 'Q2_2019')

DT_2019_Q3_df <- DT_2019_Q3_df %>% 
  mutate(quarter_year = 'Q3_2019')

DT_2019_Q4_df <- DT_2019_Q4_df %>% 
  mutate(quarter_year = 'Q4_2019')

DT_2020_Q1_df <- DT_2020_Q1_df %>% 
  mutate(quarter_year = 'Q1_2020')
```

Combine dataframes.
```{r Combine df}
DT_all <- rbind(DT_2019_Q2_df, DT_2019_Q3_df, DT_2019_Q4_df, DT_2020_Q1_df)

```

Separate DateTime column to Date and Time.
```{r Separate DateTime column}
DT_all <- tidyr::separate(DT_all, start_time, c('start_date', 'start_time'), sep = ' ')
DT_all <- tidyr::separate(DT_all, end_time, c('end_date', 'end_time'), sep = ' ')

```

Check data structure.
```{r Glimpse, message=FALSE, warning=FALSE}
glimpse(DT_all)
```

Change data structures to appropriate type and add month, day, year, day of week, and hour columns.
```{r Data structure, message=FALSE, warning=FALSE, results='hide'}
as.character(DT_all$from_station_id)
as.character(DT_all$to_station_id)
as.numeric(DT_all$tripduration)

DT_all$date <- as.Date(DT_all$start_date) #The default format is yyyy-mm-dd
DT_all$month <- format(as.Date(DT_all$date), "%m")
DT_all$day <- format(as.Date(DT_all$date), "%d")
DT_all$year <- format(as.Date(DT_all$date), "%Y")
DT_all$day_of_week <- format(as.Date(DT_all$date), "%A")

DT_all <- DT_all %>% 
  mutate(hour = hour(hms(start_time)))
```

Drop start and end date columns
```{r Remove extra date columns}
drop <- c('start_date', 'end_date')
DT_all_v2 <- DT_all[!(names(DT_all) %in% drop)]
```

Remove duplicates and NAs.
```{r Remove duplicates and NAs}
DT_all_v2 <- DT_all_v2[!duplicated(DT_all_v2$trip_id), ]
DT_all_v2 <- na.omit(DT_all_v2) 
```

Standardize usertype column.
```{r Standardize usertype}
unique(DT_all_v2$usertype)
DT_all_v2$usertype[DT_all_v2$usertype=='Subscriber'] <- 'member'
DT_all_v2$usertype[DT_all_v2$usertype=='Customer'] <- 'casual'
```

Check if stakeholder's HQ is present in the data.
```{r Check for HQ}
DT_all_v2[grep('HQ', DT_all_v2$from_station_name), ]
```

Check the minimum trip duration for negative values.
```{r Minimum trip duration}
min(DT_all_v2$tripduration)
```

Remove bike data that is from HQ or has a negative trip duration.
```{r Remove dirty data}
DT_all_v3 <- DT_all_v2[!(DT_all_v2$from_station_name == 'HQ QR' | DT_all_v2$tripduration<0),]
```

#### Analysis

##### Descriptive Statistics

<u>Below I have included the analysis path suggested by the Google Analytics R-Script:</u>

Summary of all trip durations.
```{r Summary all trip durations}
summary(DT_all_v3$tripduration)
```

Summary comparing all trip durations for days of week.
```{r Summary comparing all trip durations by days of week}
DT_all_v3$day_of_week <- ordered(DT_all_v3$day_of_week, levels=c('Sunday', 'Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday', 'Saturday'))
aggregate(DT_all_v3$tripduration~DT_all_v3$day_of_week, FUN = summary)
```

Summary comparing trip duration of user types.
```{r Summary comparing trip durations by user types}
aggregate(DT_all_v3$tripduration~DT_all_v3$usertype, FUN = summary)
```

Summary comparing trip durations of user types and days of the week.
```{r Summary comparing trip durations of user type and day of week}
aggregate(DT_all_v3$tripduration~DT_all_v3$usertype + DT_all_v3$day_of_week, FUN = summary)
```

<u>I was also interested in looking at how the user types compared at other time aggregates:</u>

Summary comparing trip durations of user types by month.
```{r Summary comparing trip durations of user type and month}
aggregate(DT_all_v3$tripduration~DT_all_v3$usertype + DT_all_v3$month, FUN = summary)
```

Summary comparing user types by year.
```{r Summary comparing trip durations of user type and year}
aggregate(DT_all_v3$tripduration~DT_all_v3$usertype + DT_all_v3$year, FUN = summary)
```

Summary comparing user types by quarter-year.
```{r Summary comparing trip durations of user type and quarter-year}
aggregate(DT_all_v3$tripduration~DT_all_v3$usertype + DT_all_v3$quarter_year, FUN = summary)
```

Compare trip durations by user type and hour of the day.
```{r Summary comparing trip durations of user type and hour}
aggregate(DT_all_v3$tripduration~DT_all_v3$usertype + DT_all_v3$hour, FUN = summary)

```

Export data for Tableau upload.
```{r Export data}
write.csv(DT_all_v3, file = 'DT_all_v3.csv')
```

